---
title: "Climatic Potential for Agricultural Productivity"
date: "Last run on `r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

```{r, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# load packages
library(sf) # geospatial analysis
library(stars) # raster processing
library(tidyverse) # data cleaning
# library(maps) used for state boundaries, but not loaded here
```

# Global productivity estimates

The Miami model of Helmut Lieth characterizes climatic potential net primary production (NPP) as a function of two limiting factors: mean annual temperature and total annual precipitation. It uses the following empirical functions to relate each variable to potential NPP, the lowest of which (i.e. the limiting factor) is taken as the final NPP estimate:

$$NPP_{precip} = 3000(1 - e^{-0.000664 P})$$
$$NPP_{temp} = \frac{3000}{1 + e^{1.315-0.119T}}$$

We'll apply this model to 30-year temperature and precipitation averages from the CHELSA dataset, a set of statistically downscaled climate data that corrects for topoclimatic influences like orographic precipitation and temperature inversions.

```{r}
npp <- read_stars(c('CHELSA_bio10_01.tif', 'CHELSA_bio10_12.tif')) %>% 
  setNames(c('temperature', 'precipitation')) %>%
  mutate(temperature = temperature / 10, # temperature is in degrees C * 10
         ### the miami model
         npp_prec = 3000 * (1 - exp(-0.000664 * precipitation)),
         npp_temp = 3000 / (1 + exp(1.315 - 0.119 * temperature)),
         npp = pmin(npp_prec, npp_temp)) %>%
  select(npp)
```

The resulting map gives us climatic potential NPP in grams of dry matter per square meter per year, with a theoretical maximum value of 3,000 $g/m^2$. This can be easily converted to expected crop yields using empirical coefficients for different crop types.

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = npp, downsample = 10) +
  scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Potential net primary production', 'Maximum terrestrial biomass given modern temperature and precipitation') +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude') +
  coord_quickmap()
```

\newpage

# Regional productivity estimates

## Southeast Mounds

Next clip the global NPP map to our regions of interest. Use the state boundaries of Alabama, Georgia, Mississippi, and Tennessee to define this study region.
```{r}
# get state boundary shapefiles
se_boundary <- maps::map('state', regions = c('alabama', 'mississippi', 'georgia', 'tennessee'), 
                    fill = TRUE, plot = FALSE) %>% 
  st_as_sf()

# subset global NPP
se_npp <- st_as_stars(npp[se_boundary])
```

```{r, echo = FALSE}
a <- ggplot() +
  geom_stars(data = se_npp) +
  geom_sf(data = se_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'US Southeast') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
a
```

\newpage

## Southwest

Use the boundaries of Arizona and New Mexico for now. There's probably a better region boundary to use on cyberSW?
```{r}
# get state boundary shapefiles
sw_boundary <- maps::map('state', regions = c('arizona', 'new mexico'), 
                    fill = TRUE, plot = FALSE) %>% 
  st_as_sf()

# subset global NPP map
sw_npp <- st_as_stars(npp[sw_boundary])
```

```{r, echo = FALSE}
b <- ggplot() +
  geom_stars(data = sw_npp) +
  geom_sf(data = sw_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'US Southwest') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
b
```

\newpage

## Central Italy 

Calculate a convex hull around all survey boundaries from Palmisano et al. 2018. Would be nice to get the original shapefile from the manuscript.
```{r}
rome <- read_sf('Palmisanoetal/Palmisanoetal/shp/surveys.shp') %>% 
  st_union() %>% 
  st_convex_hull()

rome_npp <- st_as_stars(npp[rome])
```

```{r, echo = FALSE}
c <- ggplot() +
  geom_stars(data = rome_npp) +
  geom_sf(data = rome, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'Central Italy') +
  labs(y = 'Latitude', x = 'Longitude') +  theme_bw()
c
```

\newpage

## Mixteca Alta Survey

Read in the 'border' shapefile from the Mixteca Survey folder on Dropbox. Not in lat/lon and the projection isn't clear from the file. Guessing it's originally NAD27 UTM 14N from looking at prior publications, but need to confirm!
```{r}
# this file reads as 3 linestrings
mixteca_alta <- read_sf('~/Dropbox/Settlement Persistence Project/Samples-Data/Mixteca Survey Data/border/arc.adf') %>%
  st_union() %>% # combine linestrings into 1 line
  st_polygonize() %>% # convert to polygon
  st_as_sf(crs = 26714) %>% # NAD27 UTM Zone 14N -- is this
  st_transform(4326) # transform to lat lon

mixteca_npp <- st_as_stars(npp[mixteca_alta])
```

```{r, echo = FALSE}
d <- ggplot() +
  geom_stars(data = mixteca_npp) +
  geom_sf(data = mixteca_alta, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
   ggtitle('Climatological NPP', 'Mixteca Alta survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
d
```
\newpage


## Yautepec Valley Survey

Using the "valbordr" shapefile from the yv94-96 folder on Dropbox. Assuming the projection is the same as the Mixteca data.
```{r, message = FALSE}
yautepec <- read_sf('~/Dropbox/Settlement Persistence Project/Samples-Data/Yautepec Valley Survey/yv94-96/valbordr.shp') %>% 
  st_set_crs(26714) %>% # NAD27 UTM Zone 14N -- is this
  st_transform(4326) # transform to lat lon
  
yautepec_npp <- st_as_stars(npp[yautepec])
```

```{r, echo = FALSE}
e <- ggplot() +
  geom_stars(data = yautepec_npp) +
  geom_sf(data = yautepec, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
    ggtitle('Climatological NPP', 'Yautepec Valley survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
e
```

\newpage

## Comparison

As the maps above highlight, NPP varies greatly across regions. How do their distributions look on the same scale?

```{r, echo = FALSE}
list(rome = rome_npp, sw = sw_npp, se = se_npp, mixteca = mixteca_npp, yautepec = yautepec_npp) %>%
  map_dfr(as_data_frame, .id = 'region') %>%
  remove_missing() %>%
  ggplot(aes(npp, fill = region)) +
  geom_density(alpha = .65) +
  scale_fill_brewer(palette = 'Spectral') +
  ggtitle('Regional NPP distributions') +
  theme_bw()
```

# To Do

- Crop coefficients (npp -> wheat, maize yields)
- NPP reconstructions (850 CE - present would be easiest)
- Arable land mask using slope
