---
title: "Climatic Potential for Agricultural Productivity"
date: "Last run on `r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

```{r setup, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# load packages
library(sf) # geospatial analysis
library(stars) # raster processing
library(tidyverse) # data cleaning
# library(maps) used for state boundaries, but not loaded here
```

# Global productivity estimates

The Miami model of Helmut Lieth characterizes climatic potential net primary production (NPP) as a function of two limiting factors: mean annual temperature and total annual precipitation. It uses the following empirical functions to relate each variable to potential NPP, the lowest of which (i.e. the limiting factor) is taken as the final NPP estimate:

$$NPP_{precip} = 3000(1 - e^{-0.000664 P})$$
$$NPP_{temp} = \frac{3000}{1 + e^{1.315-0.119T}}$$

We'll apply this model to 30-year temperature and precipitation averages from the CHELSA dataset, a set of statistically downscaled climate data that corrects for topoclimatic influences like orographic precipitation and temperature inversions.

```{r}
npp <- read_stars(c('CHELSA_bio10_01.tif', 'CHELSA_bio10_12.tif')) %>% 
  setNames(c('temperature', 'precipitation')) %>%
  mutate(temperature = temperature / 10, # temperature is in degrees C * 10
         ### the miami model
         npp_prec = 3000 * (1 - exp(-0.000664 * precipitation)),
         npp_temp = 3000 / (1 + exp(1.315 - 0.119 * temperature)),
         npp = pmin(npp_prec, npp_temp)) %>%
  select(npp)
```

The resulting map gives us climatic potential NPP in grams of dry matter per square meter per year, with a theoretical maximum value of 3,000 $g/m^2$. This can be easily converted to expected crop yields using empirical coefficients for different crop types.

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = npp, downsample = 10) +
  scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Potential net primary production', 'Maximum terrestrial biomass given modern temperature and precipitation') +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude')
```

\newpage

# Terrain

```{r}
arable <- c('~/Desktop/Slopes/GloSlopesCl1_30as.asc',
            '~/Desktop/Slopes/GloSlopesCl2_30as.asc', 
            '~/Desktop/Slopes/GloSlopesCl3_30as.asc', 
            '~/Desktop/Slopes/GloSlopesCl4_30as.asc') %>%
  read_stars(along = 'class') %>%
  st_set_crs(4326) %>%
  st_apply(1:2, sum, .fname = 'arable')
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = arable, downsample = 10) +
  scale_fill_viridis_c(na.value = NA, name = '% arable') +
  ggtitle('Arable land', 'Percent grid cell with slopes < 5 degrees') +
  theme_bw() +
  labs(x = 'Longitude', y = 'Latitude')
```

```{r}
c(npp, arable, along = 3)
```


# Regional productivity estimates

## Southeast Mounds

Next clip the global NPP map to our regions of interest. Use the state boundaries of Alabama, Georgia, Mississippi, and Tennessee to define this study region.
```{r}
# get state boundary shapefiles
se_boundary <- maps::map('state', regions = c('alabama', 'mississippi', 'georgia', 'tennessee'), 
                    fill = TRUE, plot = FALSE) %>% 
  st_as_sf() %>%
  st_union()

# subset global NPP and arable land
se_npp <- st_as_stars(npp[se_boundary])
se_arable <- st_as_stars(arable[se_boundary]) / 100
se_env <- se_npp * se_arable * st_area(se_arable) / 1000000 * 0.25 # the 1000000 converts from kg to kt, the 0.25 from total biomass to aboveground grain biomass
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = se_npp) +
  geom_sf(data = se_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'US Southeast') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = se_arable * 100) +
  geom_sf(data = se_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'US Southeast') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = se_env) +
  geom_sf(data = se_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'US Southeast') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```


\newpage

## Southwest

Use the boundaries of Arizona and New Mexico for now. There's probably a better region boundary to use on cyberSW?
```{r}
# get state boundary shapefiles
sw_boundary <- maps::map('state', regions = c('arizona', 'new mexico'), 
                    fill = TRUE, plot = FALSE) %>% 
  st_as_sf() %>%
  st_union()

sw_npp <- st_as_stars(npp[sw_boundary])
sw_arable <- st_as_stars(arable[sw_boundary]) / 100
sw_env <- sw_npp * sw_arable * st_area(sw_arable) / 1000000 * 0.25
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = sw_npp) +
  geom_sf(data = sw_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'US Southwest') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = sw_arable * 100) +
  geom_sf(data = sw_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'US Southwest') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = sw_env) +
  geom_sf(data = sw_boundary, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'US Southwest') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```

\newpage


## Central Italy 

Calculate a convex hull around all survey boundaries from Palmisano et al. 2018. Would be nice to get the original shapefile from the manuscript.
```{r}
rome <- read_sf('Palmisanoetal/Palmisanoetal/shp/surveys.shp') %>% 
  st_union() %>% 
  st_convex_hull()

rome_npp <- st_as_stars(npp[rome])
rome_arable <- st_as_stars(arable[rome]) / 100
rome_env <- rome_npp * rome_arable * st_area(rome_arable) / 1000000 * 0.25
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = rome_npp) +
  geom_sf(data = rome, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'Central Italy') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = rome_arable * 100) +
  geom_sf(data = rome, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'Central Italy') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = rome_env) +
  geom_sf(data = rome, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'Central Italy') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```

\newpage

## Mixteca Alta Survey

Read in the 'border' shapefile from the Mixteca Survey folder on Dropbox. Not in lat/lon and the projection isn't clear from the file. Guessing it's originally NAD27 UTM 14N from looking at prior publications, but need to confirm!
```{r}
# this file reads as 3 linestrings
mixteca_alta <- read_sf('~/Dropbox/Settlement Persistence Project/Samples-Data/Mixteca Survey Data/border/arc.adf') %>%
  st_union() %>% # combine linestrings into 1 line
  st_polygonize() %>% # convert to polygon
  st_as_sf(crs = 26714) %>% # NAD27 UTM Zone 14N -- is this
  st_transform(4326) # transform to lat lon

mixteca_npp <- st_as_stars(npp[mixteca_alta])
mixteca_arable <- st_as_stars(arable[mixteca_alta]) / 100
mixteca_env <- mixteca_npp * mixteca_arable * st_area(mixteca_arable) / 1000000 * 0.25
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = mixteca_npp) +
  geom_sf(data = mixteca_alta, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'Mixteca Alta survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = mixteca_arable * 100) +
  geom_sf(data = mixteca_alta, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'Mixteca Alta survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = mixteca_env) +
  geom_sf(data = mixteca_alta, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'Mixteca Alta survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```
\newpage


## Yautepec Valley Survey

Using the "valbordr" shapefile from the yv94-96 folder on Dropbox. Assuming the projection is the same as the Mixteca data.
```{r, message = FALSE}
yautepec <- read_sf('~/Dropbox/Settlement Persistence Project/Samples-Data/Yautepec Valley Survey Data/GIS Files yv94-96/valbordr.shp') %>% 
  st_set_crs(26714) %>% # NAD27 UTM Zone 14N -- is this
  st_transform(4326) # transform to lat lon

yautepec_npp <- st_as_stars(npp[yautepec])
yautepec_arable <- st_as_stars(arable[yautepec]) / 100
yautepec_env <- yautepec_npp * yautepec_arable * st_area(yautepec_arable) / 1000000 * 0.25
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = yautepec_npp) +
  geom_sf(data = yautepec, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'Yautepec Valley survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data =yautepec_arable * 100) +
  geom_sf(data = yautepec, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'Yautepec Valley survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = yautepec_env) +
  geom_sf(data = yautepec, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'Yautepec Valley survey') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```

\newpage


## Fertile Crescent

Load the Fragile Crescent project boundary shapefile.
```{r}
crescent <- read_sf('~/Dropbox/Settlement Persistence Project/Samples-Data/Near East Data/GIS/FertileCrescent_region.shp')

crescent_npp <- st_as_stars(npp[crescent])
crescent_arable <- st_as_stars(arable[crescent]) / 100
crescent_env <- crescent_npp * crescent_arable * st_area(crescent_arable) / 1000000 * 0.25
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = crescent_npp) +
  geom_sf(data = crescent, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'Fertile Crescent') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data =crescent_arable * 100) +
  geom_sf(data = crescent, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'Fertile Crescent') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = crescent_env) +
  geom_sf(data = crescent, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'Fertile Crescent') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```
\newpage


## Basin of Mexico

Read the Basin of Mexico hydrological boundary shapefile.
```{r}
bom <- read_sf('a2606.shp') %>% 
  st_union() %>%
  st_transform(st_crs(npp))

bom_npp <- st_as_stars(npp[bom])
bom_arable <- st_as_stars(arable[bom]) / 100
bom_env <- bom_npp * bom_arable * st_area(bom_arable) / 1000000 * 0.25
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = bom_npp) +
  geom_sf(data = bom, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'Basin of Mexico') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data =bom_arable * 100) +
  geom_sf(data = bom, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'Basin of Mexico') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = bom_env) +
  geom_sf(data = bom, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'Basin of Mexico') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```

\newpage


## Santa Valley

```{r}
santa <- read_sf('santa_valley.shp') %>% 
  st_union() %>%
  st_transform(st_crs(npp))

santa_npp <- st_as_stars(npp[santa])
santa_arable <- st_as_stars(arable[santa]) / 100
santa_env <- santa_npp * santa_arable * st_area(santa_arable) / 1000000 * 0.25
```

```{r, echo = FALSE}
ggplot() +
  geom_stars(data = santa_npp) +
  geom_sf(data = santa, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = expression(g/m^2)) +
  ggtitle('Climatological NPP', 'Santa Valley') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data =santa_arable * 100) +
  geom_sf(data = santa, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = '%') +
  ggtitle('Arable land', 'Santa Valley') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()

ggplot() +
  geom_stars(data = santa_env) +
  geom_sf(data = santa, fill = NA, size = 1) +
    scale_fill_viridis_c(na.value = NA, name = 'Kilotons grain') +
  ggtitle('Environmental potential', 'Santa Valley') +
  labs(y = 'Latitude', x = 'Longitude') +
  theme_bw()
```

\newpage

## Comparison

As the maps above highlight, NPP varies greatly across regions. How do their distributions look on the same scale?

```{r, echo = FALSE}
list(rome = rome_npp, sw = sw_npp, se = se_npp, yautepec = yautepec_npp, crescent = crescent_npp, bom = bom_npp, santa = santa_npp) %>%
  map_dfr(as_data_frame, .id = 'region') %>%
  remove_missing() %>%
  ggplot(aes(npp, fill = reorder(region, npp, FUN = median))) +
  geom_density(alpha = .65) +
  scale_fill_brewer(palette = 'Spectral', name = 'Region') +
  ggtitle('Regional NPP distributions') +
  theme_bw()
```

```{r}
library(ggridges)
list(rome = rome_npp, sw = sw_npp, se = se_npp, yautepec = yautepec_npp, crescent = crescent_npp, bom = bom_npp, santa = santa_npp) %>%
  map_dfr(as_data_frame, .id = 'region') %>%
  remove_missing() %>%
  ggplot(aes(npp, reorder(region, -npp, FUN = median), fill = reorder(region, npp, FUN = median))) +
  geom_density_ridges(alpha = .65) +
  scale_fill_brewer(palette = 'Spectral', name = 'Region') +
  ggtitle('Regional NPP distributions') +
  labs(y = '') +
  theme_bw()
```

```{r, echo = FALSE}
list(rome = rome_env, sw = sw_env, se = se_env, yautepec = yautepec_env, crescent = crescent_env, bom = bom_env, santa = santa_env) %>%
  map_dfr(as_data_frame, .id = 'region') %>%
  remove_missing() %>%
   ggplot(aes(as.numeric(npp), reorder(region, -as.numeric(npp), FUN = median), fill = reorder(region, as.numeric(npp), FUN = median))) +
  geom_density_ridges(alpha = .65) +
  scale_fill_brewer(palette = 'Spectral', name = 'Region') +
  ggtitle('Regional NPP distributions with arable land mask') +
    labs(y = '', y = 'npp') +
  theme_bw()
```
```{r}
list(rome = rome_npp, sw = sw_npp, se = se_npp, yautepec = yautepec_npp, crescent = crescent_npp, bom = bom_npp, santa = santa_npp) %>%
  map_dfr(as_data_frame, .id = 'region') %>%
  remove_missing() %>%
  ggplot(aes(npp, reorder(region, -npp, FUN = median), fill = stat(x))) +
    geom_density_ridges_gradient(alpha = .65) +
  scale_fill_viridis_c(name = 'Region') +
  ggtitle('Regional NPP distributions') +
  labs(y = '') +
  theme_bw()
```


# ArchaeGLOBE Regions
What if instead we interpret the settlement data as *samples* from a larger distribution, and look at more "objective" regions

```{r}
ag_simp <- read_sf('../ArchaeoGLOBE/analysis/data/raw-data/ArchaeoGLOBE_Simplified_Regions/ArchaeoGLOBE_Simplified_Regions.shp') %>% filter(Archaeo_RG %in% c('Southwestern USA', 'Southeastern USA', 'Central Mexico', 'Syria, Cyprus, and the Levant', 'Iraq', 'Gran Chaco', 'Southern Europe')) %>%
  st_transform(st_crs(npp))
```

```{r}
ag <- read_sf('../ArchaeoGLOBE/analysis/data/raw-data/ArchaeoGLOBE_Regions/ArchaeGLOBE_Regions.shp') %>% filter(Archaeo_RG %in% c('Southwestern USA', 'Southeastern USA', 'Central Mexico', 'Syria, Cyprus, and the Levant', 'Iraq', 'Gran Chaco', 'Southern Europe')) %>%
  st_transform(st_crs(npp))
```

```{r}
ag_npp <- ag %>%
  select(Archaeo_RG) %>%
    mutate(bb = split(., 1:7) %>% map(~st_crop(npp, .x))) %>%
  st_drop_geometry() %>%
  mutate(dat = map(bb, ~st_as_stars(.) %>% as_tibble)) %>%
  select(-bb) %>%
  unnest(dat) %>%
  remove_missing()
  

ggplot(ag_npp, aes(npp, fill = Archaeo_RG)) +
  geom_density(alpha = .65) +
  scale_fill_brewer(palette = 'Spectral') +
  ggtitle('Regional NPP distributions') +
  theme_bw()
```
```{r}
ag %>% filter(Archaeo_RG == 'Gran Chaco')
```


# To Do

- Crop coefficients (npp -> wheat/maize yields)
- NPP reconstructions (850 CE - present would be easiest)
